A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2016-11-08 Draft14.tex
small edits
small changes

In this paper, we propose a bridge between continuous-time limits and discrete-time algorithms.  Our method takes, as its primary object of study a general Lyapunov function for momentum methods in continuous time.  Through a diverse set of examples, we demonstrate how the proofs of momentum methods can be understood as bounding discretization errors of the Lyapunov function when moving to discrete time.  In particular, we show how the discretization of the associated continuous-time ODE needs to be performed in such a way that the Lyapunov function remains valid when transitioning to discrete time.
Using this technique, we provide a simple, direct proof of the convergence of Nesterov's method for strongly convex functions in Euclidean spaces.  Our proof makes clear why only certain parameterizations of the continuous ODEs lead to valid discrete-time methods.  Moreover, we explain the need for the extragradient step inside Nesterov's method which does not appear in Polyak's method.  
Finally, we show there is a striking equivalence between estimate sequences and Lyapunov functions.  In continuous and discrete time, estimate sequences can be derived directly from the Lyapunov function and vice versa.  We show that the associated continuous-time estimate sequence can also be directly discretized to give standard estimate-sequence based proofs of momentum methods.  This clarifies how Nesterov's ``algebraic trick'' is closely related to invariant reduction proofs that are common in algorithm analysis.

@@

Until recently, this ODE perspective on optimization algorithms was largely abandoned in favor of Nesterov's estimate sequence framework.  However, there has been a recent resurgence in interest in ODE-based analyses.  Su, Boyd and Cand\`es~\cite{SuBoydCandes14}, showed that the continuous-time limit of Nesterov's accelerated gradient descent method corresponds to the following second-order ODE, 
 \begin{equation}\label{Eq:SuBoydCandes}
 \ddot X_t + \frac{r}{t} \dot X_t + \nabla f(X_t) = 0,\,\,\,\,r \geq 3.
 \end{equation}
This is equivalent to~\eqref{eq:heavy-ball-ode}, where the dynamics model an object with unit mass and the damping term $\gamma_t = r\log t$ is logarithmic.  They also established a convergence rate of $O(1/t^2)$ for this continuous-time system via a Lyapunov function argument. This matches the $O(1/\epsilon k^2)$ convergence rate of accelerated gradient descent, given the discretization $t= \sqrt\epsilon k$. Thus, the convergence rate is maintained in passing from continuous time to discrete time. Two things remain unclear from their analysis however:  they introduce an additional sequence $\{y_k\}_{k=1}^\infty$ while discretizing the ODE~\cite[Sec 2]{Acceleration},  which does not correspond to a straightforward discretization technique; and they do not demonstrate what connection---if any---there is between the Lypaunov argument used to analyze the continuous-time dynamics and the technique of estimate sequences used to analyze the discrete-time algorithm.

@@

0 \leq \langle \nabla f(X_t), x-  Z_t\rangle, \quad \quad \forall x \in \X 

@@

>>>>

In this paper, we propose a bridge between continuous-time limits and discrete-time algorithms.  Our method takes, as its primary object of study a general Lyapunov function for momentum methods in continuous time.  Through a diverse set of examples, we demonstrate how the proofs of momentum methods can be understood as bounding discretization errors of the Lyapunov function when moving to discrete time.  In particular, we show how the discretization of the associated continuous-time ODE needs to be performed in such a way that the Lyapunov function remains valid when transitioning to discrete time.
Using this technique, we provide a simple, direct proof of the convergence of Nesterov's method for strongly convex functions in Euclidean spaces.  Our proof makes clear why only certain parameterizations of the continuous ODEs lead to valid discrete-time methods.  Moreover, we provide new perspectives on the need for the extragradient step inside Nesterov's method which does not appear in Polyak's method.  
Finally, we show there is a striking equivalence between estimate sequences and Lyapunov functions.  In continuous and discrete time, estimate sequences can be derived directly from the Lyapunov function and vice versa.  We show that the associated continuous-time estimate sequence can also be directly discretized to give standard estimate-sequence based proofs of momentum methods.  This clarifies how Nesterov's ``algebraic trick'' is closely related to invariant reduction proofs that are common in algorithm analysis.

@@

Until recently, this ODE perspective on optimization algorithms was largely abandoned in favor of Nesterov's estimate sequence framework.  However, there has been a recent resurgence in interest in ODE-based analyses.  Su, Boyd and Cand\`es~\cite{SuBoydCandes14}, showed that the continuous-time limit of Nesterov's accelerated gradient descent method corresponds to the following second-order ODE, 
 \begin{equation}\label{Eq:SuBoydCandes}
 \ddot X_t + \frac{r}{t} \dot X_t + \nabla f(X_t) = 0,\,\,\,\,r \geq 3.
 \end{equation}
This is equivalent to~\eqref{eq:heavy-ball-ode}, where the dynamics model an object with unit mass and the damping term is logarithmic, $\gamma_t = r\log t$.  They also established a convergence rate of $O(1/t^2)$ for this continuous-time system via a Lyapunov function argument. This matches the $O(1/\epsilon k^2)$ convergence rate of accelerated gradient descent, given the discretization $t= \sqrt\epsilon k$. Thus, the convergence rate is maintained in passing from continuous time to discrete time. Two things remain unclear from their analysis however:  they introduce an additional sequence $\{y_k\}_{k=1}^\infty$ while discretizing the ODE~\cite[Sec 2]{Acceleration},  which does not correspond to a straightforward discretization technique; and they do not demonstrate what connection---if any---there is between the Lypaunov argument used to analyze the continuous-time dynamics and the technique of estimate sequences used to analyze the discrete-time algorithm.

@@

0 \leq \langle \nabla f(X_t), x-  Z_t\rangle,

@@

 $\forall x \in \X $, where 
